Learning STRIPS action models with classical planning

### 2017-10-05 doc/icaps-18/strips-learning.tex
updates to page 2 icaps18 paper

\subsection{Classical Planning}
We use $F$ to denote the set of {\em fluents} (propositional variables) describing a state. A {\em literal} $l$ is a valuation of a fluent $f\in F$, i.e.~$l=f$ or $l=\neg f$. A set of literals $L$ represents a partial assignment of values to fluents (WLOG we assume that $L$ does not assign conflicting values to any fluent). We use $\mathcal{L}(F)$ to denote the set of all literal sets on $F$, i.e.~all partial assignments of values to fluents. A {\em state} $s$ is then a total assignment of values to fluents, i.e. $|s|=|F|$, so the size of the state space $2^{|F|}$. Explicitly including negative literals $\neg f$ in states simplifies subsequent definitions, but we often abuse notation by defining a state $s$ only in terms of the fluents that are true in $s$, as is common in \strips\ planning.

@@

In this work we assume that fluents $F$ are instantiated from predicates, as in PDDL~\cite{mcdermott1998pddl,fox2003pddl2}. There exists a set of predicates $\Psi$, each $p\in\Psi$ with an argument list of arity $ar(p)$. Given a set of objects $\Omega$, the set of fluents $F$ is then induced by assigning objects in $\Omega$ to the arguments of predicates in $\Psi$, i.e.~$F=\{p(\omega):p\in\Psi,\omega\in\Omega^{ar(p)}\}$ s.t. $\Omega^n$ is the $n$-th Cartesian power of $\Omega$.

Likewise, we assume that actions in $A$ are instantiated from an STRIPS operator schema. Let $\Omega_v=\{v_1,\ldots,v_v\}$, $\Omega_v\cap\Omega=\emptyset$ be a new set of objects representing {\em variable names} and that is defined by the action with the maximum arity in a planning frame. For instance, in a three-block blocksworld then $\Omega=\{block_1, block_2, block_3\}$ and $\Omega_v=\{v_1, v_2\}$ since the actions instantiated from the operators {\small\tt stack} (see Figure~\ref{fig:stack}) and {\small\tt unstack} have two parameters. Let us define a new set of fluents $F_{v}$ that results instantiating $\Psi$ but using only the objects in $\Omega_v$, e.g. for the blocksworld example $F_v$={\small\tt\{handempty, holding($v_1$), holding($v_2$), clear($v_1$), clear($v_2$), ontable($v_1$), ontable($v_2$), on($v_1,v_1$),on($v_1,v_2$), on($v_2,v_1$),on($v_2,v_2$)\}}.

\begin{figure}[hbt]
\begin{footnotesize}
\begin{scriptsize}
\begin{verbatim}
  (:action stack
   :parameters (?x1 ?x2)
   :precondition (and (holding ?x1) 
                      (clear ?x2))
   :effect (and (not (holding ?x1))
                (not (clear ?x2))
                (clear ?x1)
                (handempty)
                (on ?x1 ?x2)))
\end{verbatim}
\end{scriptsize}
\end{footnotesize}
 \caption{\small Example of the {\em stack} planning operator schema from the blocksworld as represented in PDDL.}
\label{fig:stack}
\end{figure}

We are now ready to define a STRIPS operator schema as a tuple $\xi=\tup{head(\xi),pre(\xi),add(\xi),del(\xi)}$ where:
\begin{itemize}
\item $head(\xi)=\tup{name(\xi),pars(\xi)}$, represents an operator {\em header} defined by its corresponding action name and a list of variables, $pars(\xi)\in\Omega_v^{ar(\xi)}$. Learning the action model for the blocksworld requires these four operator headers, $\Xi_{blocks}=${\small\tt\{pickup($v_1$), putdown($v_1$), stack($v_1,v_2$), unstack($v_1,v_2$)\}}.
\item The preconditions $pre(\xi)\subseteq F_v$, the positive effects $add(\xi)\subseteq F_v$, and the negative effects $del(\xi)\subseteq F_v$ such that, $del(\xi)\subseteq pre(\xi)$, $del(\xi)\cap add(\xi)=\emptyset$ and $pre(\xi)\cap add(\xi)=\emptyset$.
\end{itemize}

\subsection{Classical Planning with Conditional Effects}
Our approach for leaning STRIPS operator schema is compiling this leaning task into a classical planning task with conditional effects. Conditional effects make it possible to refer to an action while their precise effects depend on the current state.

@@

\section{Learning classical planning action models}
Learning STRIPS action models from fully available input knowledge, i.e. a set of plans where the {\em pre-} and {\em post-states} of every action in a plan are available, is straightforward. In this case, the operators schema are derived lifting the literals that change between the pre and post-state of the corresponding action executions. Preconditions are derived lifting the minimal set of literals appearing in all the pre-states that correspond to the same operator.

@@

\subsection{Learning from labeled plans}
This learning task is formalized as $\Lambda=\tup{\Psi,\Pi,\Sigma}$: 
\begin{itemize}
\item $\Psi$ the set of predicates that define the abstract state space of a given planning domain. This set includes the predicates for defining the headers of the operators schema.
\item $\Pi=\{\pi_1,\ldots,\pi_{\tau}\}$ is the given set of example plans,
\item $\Sigma=\{\sigma_1,,\ldots,\sigma_{\tau}\}$ is a set of labels s.t. each plan $\pi_t$, {\small $1\leq t\leq \tau$}, has an associated label $\sigma_t=(s_0^t,s_{n+1}^t)$ where $s_{n+1}^t$ is the state resulting from executing $\pi_t$ starting from the state $s_0^t$. 
\end{itemize}

@@

\subsection{Learning from states}
Here we reduce the amount of input knowledge provided to the previous learning task. Now $\Pi=\{\pi_1,\ldots,\pi_t\}$ is replaced by  $\Pi'=\{|\pi_1|,\ldots,|\pi_t|\}$ that does not contain a set of plans but the number of actions of each plan, so the learning task is redefined as $\Lambda'=\tup{\Psi,\Pi',\Sigma}$. While the previous task can correspond to watching an agent acting in the world, this learnig tasks can be understood as watching only the results of its actions executions knowing the number of diferent actions performed by the agent.

Finally, we can go a step further and redefine a third learning task $\Lambda''=\tup{\Psi,\Sigma}$ that corresponds to watching only the results of the plan executions. Now a solution to the $\Lambda''$ learning task is a set of operator schema $\Xi$ that is compliant only with the predicates in $\Psi$, and the given set of initial and final states.

\section{Learning action models from example states using a classical planner}
Our approach for addressing $\Lambda''=\tup{\Psi,\Sigma}$ is compiling this learning task into a classical planning task $P_{\Lambda}$. The intuition behind the compilation is that a solution to $P_{\Lambda}$ is a sequence of actions that first, programs the action model (i.e. the $pre(\xi)$, $del(\xi)$ and $add(\xi)$ for each $\xi\in\Xi$) and then, validates the programmed action model in the given set of labels $\Sigma$, one after the other.

To formalize the compilation we define $t$ classical planning instances $P_1=\tup{F,A,I_1,G_1},\ldots,P_t=\tup{F,A,I_t,G_t}$, that belong to the same planning frame (share the same fluents and actions and differ only in the initial state and goals). The set of fluents $F$ is built instantiating the predicates in $\Psi$ with the objects in $\Omega=\{o|o\in s_i\cup s_i', {\small 1\leq i\leq t}\}$, the set of objects that appear in the fluents $f\in F$ used to define the states in $\Sigma$. The set of actions is empty $A=\emptyset$. Finally the initial state $I_i$, {\small $1\leq i\leq t$}, is given by the state $s_i\in \sigma_i$ while goals $G_i$ are defined by the state $s_i'\in \sigma_i$.

Now we are ready to define the compilation for learning action models using an off-the-shelf classical planner. Given a learning task $\Lambda'=\tup{\Psi,\Sigma}$ the compilation outputs a classical planning task $P_{\Lambda}=\tup{F_{\Lambda},A_{\Lambda},I_{\Lambda},G_{\Lambda}}$ where:
\begin{itemize}
\item $F_{\Lambda}$ extends $F$ with:
\begin{itemize}
\item Fluents representing the programmed action model: $pre_f(\xi)$, $del_f(\xi)$ and $add_f(\xi)$ for every $f\in F_v$ and $\xi \in \Xi$.
\item Fluents $\{test_i\}_{1\leq i\leq t}$, indicating the example where the programmed model is currently being validated.
\item Fluents $prog1$, $prog2$ and $exec$ indicating whether the solution is programming the preconditions of the action schema, the effects, or it started validating the programmed action models.
\end{itemize}
\item $I_{\Lambda}$, contains the fluents from $F$ that encode the initial state $s_1\in P_1$, every fluent $pre_f(\xi)\in F_{\Lambda}$ and $prog1$. 
\item $G_{\Lambda}=\{test_i\}$,{\small $1\leq i\leq t$}, indicates that the programmed action model is validated in all the examples in $\Sigma$.
\item $A_{\Lambda}$ replaces the actions in $A$ with actions of three types:
\begin{enumerate}
\item Actions for programming:

>>>>

\subsection{Classical planning}
We use $F$ to denote the set of {\em fluents} (propositional variables) describing a state. A {\em literal} $l$ is a valuation of a fluent $f\in F$, i.e.~$l=f$ or $l=\neg f$. A set of literals $L$ represents a partial assignment of values to fluents (WLOG we assume that $L$ does not assign conflicting values to any fluent). We use $\mathcal{L}(F)$ to denote the set of all literal sets on $F$, i.e.~all partial assignments of values to fluents. A {\em state} $s$ is then a total assignment of values to fluents, i.e. $|s|=|F|$, so the size of the state space $2^{|F|}$. Explicitly including negative literals $\neg f$ in states simplifies subsequent definitions, but we often abuse notation by defining a state $s$ only in terms of the fluents that are true in $s$, as is common in \strips\ planning.

@@

In this work we assume that the fluents in $F$ are instantiated from predicates, as in PDDL~\cite{mcdermott1998pddl,fox2003pddl2}. There exists a set of predicates $\Psi$, each $p\in\Psi$ with an argument list of arity $ar(p)$. Given a set of objects $\Omega$, the set of fluents $F$ is then induced by assigning objects in $\Omega$ to the arguments of predicates in $\Psi$, i.e.~$F=\{p(\omega):p\in\Psi,\omega\in\Omega^{ar(p)}\}$ s.t. $\Omega^k$ is the $k$-th Cartesian power of $\Omega$.

Likewise, we assume that each action in $A$ is instantiated from an STRIPS operator schema. Figure~\ref{fig:stack} shows the {\em stack} operator schema for a STRIPS blocksworld represented in PDDL.
\begin{figure}[hbt]
\begin{footnotesize}
\begin{scriptsize}
\begin{verbatim}
(:action stack
  :parameters (?x1 ?x2)
  :precondition (and (holding ?x1) (clear ?x2))
  :effect (and (not (holding ?x1)) (not (clear ?x2))
               (clear ?x1) (handempty) (on ?x1 ?x2)))
\end{verbatim}
\end{scriptsize}
\end{footnotesize}
 \caption{\small Example of a {\em stack} operator schema for a STRIPS blocksworld represented in PDDL.}
\label{fig:stack}
\end{figure}

Let $\Omega_v$ be a new set of objects, $\Omega\cap\Omega_v=\emptyset$, that represent {\em variable names}. $|\Omega_v|$ is given by the action with the maximum arity in a planning frame. For instance, in a three-block blocksworld $\Omega=\{block_1, block_2, block_3\}$ and $\Omega_v=\{v_1, v_2\}$ because the operators {\small\tt stack} and {\small\tt unstack} are the ones with the maximum arity (two parameters each).

Let us define a new set of fluents $F_{v}$ that results instantiating $\Psi$ but using only the {\em variable objects} $\Omega_v$. In the blocksworld $F_v$={\small\tt\{handempty, holding($v_1$), holding($v_2$), clear($v_1$), clear($v_2$), ontable($v_1$), ontable($v_2$), on($v_1,v_1$),on($v_1,v_2$), on($v_2,v_1$),on($v_2,v_2$)\}}.

We are now ready to define a STRIPS operator schema as a tuple $\xi=\tup{head(\xi),pre(\xi),add(\xi),del(\xi)}$:
\begin{itemize}
\item $head(\xi)=\tup{name(\xi),pars(\xi)}$, represents an operator {\em header} defined by its corresponding action name and a list of variables, $pars(\xi)\in\Omega_v^{ar(\xi)}$. The headers for the blocksworld operators are {\small\tt pickup($v_1$), putdown($v_1$), stack($v_1,v_2$)} and {\small\tt unstack($v_1,v_2$)}.
\item The preconditions $pre(\xi)\subseteq F_v$, the positive effects $add(\xi)\subseteq F_v$, and the negative effects $del(\xi)\subseteq F_v$ such that, $del(\xi)\subseteq pre(\xi)$, $del(\xi)\cap add(\xi)=\emptyset$ and $pre(\xi)\cap add(\xi)=\emptyset$.
\end{itemize}

\subsection{Classical planning with conditional effects}
Our approach for leaning STRIPS action models is compiling this leaning task into a classical planning task with conditional effects. We use conditional effects because they allow us to compactly define actions whose effects depend on the current state. Many classical planners cope with conditional effects without compiling them away. In fact, the support of PDDL conditional effects was a requirement for participating at IPC-2014~\cite{vallati:IPC:AIM2015}.

@@

\section{Learning STRIPS action models}
Learning STRIPS action models from fully available input knowledge, i.e. a set of plans where the {\em pre-} and {\em post-states} of every action in a plan are available, is straightforward. In this case, the operators schema are derived lifting the literals that change between the pre and post-state of the corresponding action executions. Likewise, preconditions are derived lifting the minimal set of literals appearing in all the pre-states that correspond to the same operator.

@@

\subsection{Learning from labeled plans}
This learning task is formalized as $\Lambda=\tup{\Psi,\Pi,\Sigma}$: 
\begin{itemize}
\item $\Psi$, the set of predicates that define the abstract state space of a given planning domain. This set includes the predicates for defining the headers of the operators schema.
\item $\Pi=\{\pi_1,\ldots,\pi_{\tau}\}$, the given set of example plans.
\item $\Sigma=\{\sigma_1,,\ldots,\sigma_{\tau}\}$, a set of labels s.t. each plan $\pi_t$, {\small $1\leq t\leq \tau$}, has a label $\sigma_t=(s_0^t,s_{n}^t)$ where $s_{n}^t$ is the state resulting from executing $\pi_t$ starting from the state $s_0^t$. 
\end{itemize}

@@

\subsection{Learning from initial/final states}
Here we reduce the amount of input knowledge provided to the learning task. Now $\Pi=\{\pi_1,\ldots,\pi_{\tau}\}$ is replaced by  $\Pi'=\{|\pi_1|,\ldots,|\pi_{\tau}|\}$ i.e. $\Pi'$ that does not contain a set of plans but the number of actions of each plan, so the learning task is redefined as $\Lambda'=\tup{\Psi,\Pi',\Sigma}$. While the previous learning task, $\Lambda$, corresponds to watching an agent acting in the world, this new learning task $\Lambda'$ can be understood as watching only the results of its actions executions knowing the number of diferent actions performed by the agent.

Finally, we can go a step further and redefine a third learning task $\Lambda''=\tup{\Psi,\Sigma}$ that corresponds to watching only the results of the plan executions. In this case a solution to the $\Lambda''$ learning task is a set of operator schema $\Xi$ that is compliant only with the predicates in $\Psi$, and the given set of initial and final states $\Sigma$.

In these two cases, a solution must not only synthesize the action models but also the actions that produced the given labels (this information is no longer given in the learning examples).

\section{Learning STRIPS action models with classical planning}
Our approach for addressing a learning task $\Lambda$, $\Lambda'$ or $\Lambda''$, is to compile it into a classical planning task $P_{\Lambda}$, $P_{\Lambda'}$ or $P_{\Lambda''}$. The intuition behind these compilations is that a solution to the resulting classical planning task is a sequence of actions that:
\begin{enumerate}
\item Programs the action model i.e. determines the literals in the sets $pre(\xi)$, $del(\xi)$ and $add(\xi)$ for each $\xi\in\Xi$.
\item Validates the programmed action model in the given set of labels $\Sigma=\{\sigma_1,\ldots,\sigma_{\tau}\}$ i.e., for every {\small $1\leq t\leq \tau$}, uses the programmed action model $\Xi$ to produce a final state $s_{n}^t$ starting from its corresponding initial state $s_0^t$.
\end{enumerate}

To formalize these compilations we define {\small $1\leq t\leq \tau$} classical planning instances $P_t=\tup{F,A,I_t,G_t}$, that belong to the same planning frame (share the same fluents and actions and differ only in the initial state and goals). The set of fluents $F$ is built instantiating the predicates in $\Psi$ with the objects appearing in the labels. Formally $\Omega=\{o|o\in s_0^t\cup s_n^t\cup\pi_t, {\small 1\leq t\leq \tau}\}$. The set of actions is empty $A=\emptyset$, is the aim of the learning tasks adressed in the paper. Finally the initial state $I_t$ is given by the state $s_0^t\in \sigma_t$ while goals $G_t$ are defined by the state $s_n^t\in \sigma_t$.

Now we are ready to define the compilation for learning STRIPS action models using classical planning with conditional effects. Given a learning task $\Lambda''=\tup{\Psi,\Sigma}$ the compilation outputs a classical planning task $P_{\Lambda''}=\tup{F_{\Lambda},A_{\Lambda},I_{\Lambda},G_{\Lambda}}$ where:
\begin{itemize}
\item $F_{\Lambda}$ extends $F$ with:
\begin{itemize}
\item Fluents representing the programmed action model: $pre_f(\xi)$, $del_f(\xi)$ and $add_f(\xi)$ for every $f\in F_v$ and $\xi \in \Xi$.
\item Fluents $\{test_t\}_{1\leq t\leq \tau}$, indicating the example where the programmed model is currently being validated.
\item Fluents $mode_{pre}$, $mode_{eff}$ and $mode_{val}$ indicating whether currently the preconditions of the action schema are being programmed, the effects of the action schema are being programmed or the programmed action models are being validated.
\end{itemize}
\item $I_{\Lambda}$, contains the fluents from $F$ that encode $s_0^1$, every fluent $pre_f(\xi)\in F_{\Lambda}$ and $mode_{pre}$. 
\item $G_{\Lambda}=\{test_t\}$,{\small $1\leq t\leq \tau$}, indicates that the programmed action model is validated in all the learning examples.
\item $A_{\Lambda}$ contains actions of three types:
\begin{enumerate}
\item Actions for programming:
